\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-04 paper/ms.tex
comparisons

There are many other methods of scaling GP models to large datasets and in
this section we draw comparisons between \celerite\ and other popular methods.
Scalable GP methods tend to fall into two categories: approximate and
restrictive.
\celerite\ falls into the latter category because, while the method is exact,
it requires a specific choice of stationary kernel function and it can only be
used in one-dimension.

As discussed previously, the \celerite\ method is based upon a method
developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
In the case where the kernel function has a single term of the form given in
\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
and the Rybicki \& Press method.
Their method does not require solving the extended matrix because the matrix
inverse can be calculated analytically, but the method is made slightly more
complicated when white noise is included.
We implement their full method and find empirically that it is more
computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
but we note that the requirement of a single, real, exponential kernel is
extremely restrictive and this model cannot describe quasi-periodic
variability.

@@

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
is, in practice, somewhat faster and the memory requirements are smaller.
One benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.

>>>>

There are other methods of scaling GP models to large datasets and in this
section we draw comparisons between \celerite\ and other popular methods.
Scalable GP methods tend to fall into two categories: approximate and
restrictive.
\celerite\ falls into the latter category because, while the method is exact,
it requires a specific choice of stationary kernel function and it can only be
used in one-dimension.

\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}

\response{
The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
is, in practice, somewhat faster and the memory requirements are smaller.
One benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.
}

\subsection{Other scalable methods}

%As discussed previously, the \celerite\ method is based upon a method
%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
%In the case where the kernel function has a single term of the form given in
%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
%and the Rybicki \& Press method.
%Their method does not require solving the extended matrix because the matrix
%inverse can be calculated analytically, but the method is made slightly more
%complicated when white noise is included.
%We implement their full method and find empirically that it is more
%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
%but we note that the requirement of a single, real, exponential kernel is
%extremely restrictive and this model cannot describe quasi-periodic
%variability.

@@
